Generating Images From Captions\\ With Attention

### 2015-10-30 iclr-paper-new.tex
u
update

Unlike the original DRAW network where latent variables are independent unit Gaussians $\mathcal{N}(0, I)$, the latent variables in the proposed alignDRAW model have their mean and variance depend on the previous recurrent hidden states $\hdec_{t-1}$, except for $\lat_1 \sim \prior(\Lat_1) = \mathcal{N}(0, I)$.  Namely, the mean and variance of prior distribution $\prior(\Lat_t)$ is parameterized by: 
\begin{align}
\mu_{t}(\hdec_{t-1}) &= tanh(W_{\mu}\hdec_{t-1}),\\
\sigma_{t}(\hdec_{t-1}) &= {exp}\big(tanh(W_{\sigma}\hdec_{t-1})\big),
\end{align}
where $W_{\mu}, W_{\sigma}$ are learnt model parameters. It has been shown in \citep{bachman_sdm} that model performance is improved by including dependencies between latent variables. Formally, the image is generated by iteratively computing the following equations for $t=1,...,T$ (see Figure~\ref{fig:figmodel}):
\begin{align}
\label{eq:x_hat}
\lat_t &\sim \prior(\Lat_t|\Lat_{1:t-1}) = \mathcal{N}\big(\mu_{t}(\hdec_{t-1}), \sigma_{t}(\hdec_{t-1})\big),\\
s_{t} &= align(\hdec_{t-1}, \hlangall), \\
\hdec_t &= \decoder(\hdec_{t-1}, [z_t, s_{t}]), \\
\canv_t &= \canv_{t-1} + \writeop(\hdec_t), 
\label{eq:write}
%\canv_t &= \canv_{t-1} + \writeop(\hdec_t)
\end{align}
where $\mathit{LSTM}$ is a standard \textit{Long Short-Term Memory} architecture \citep{hochreiter_lstm} with \textit{forget} gates \citep{gers_forget}, $[v_1, v_2]$ is the concatenation of the vectors $v_1$ and $v_2$ into the single vector, and $\canv_0$, $\hdec_0$ are initialized to the the learned biases. The $\writeop$ operator applies an array of Gaussian filters on the patch that places it on the canvas with smoothly varying location and zoom (same as in \citep{gregor_draw}). The $align$ function is used to compute the alignment between the input caption and intermediate image generative steps \citep{bahdanau_mt}. Given the caption representation from the language model, $\hlangall = [\hlang_{1}, \hlang_{2}, ..., \hlang_{N}]$, the $align$ operator outputs a dynamic sentence representation $s_t$ at each step through a weighted sum using alignment probabilities $\alpha_{1...N}^{t}$:
%\begin{align}
%\vspace{-0.1in}
$s_t=align(\hdec_{t-1}, \hlangall) = \alpha_{1}^{t}\hlang_{1} + \alpha_{2}^{t}\hlang_{2} + ... + \alpha_{N}^{t}\hlang_{N}.$
%\end{align}
The corresponding alignment probabilities $\alpha_{1...N}^{t}$ at each step are obtained using the caption representation $\hlangall$ and the hidden state of the generative model $\hdec_{t}$, by first calculating energies using a $tanh$ perceptron scaled by vector $v$ and then normalizing them, as in \citep{bahdanau_mt}:
%\begin{align}
%\vspace{-0.1in}
$\alpha_{k}^{t}=\frac{exp(v^{T}tanh(U\hlang_{k} + W\hdec_{t} + b)}{\sum_{j=1}^{N}exp(v^{T}tanh(U\hlang_{j} + W\hdec_{t} + b)}.$
%\end{align}
The final canvas $\canv_T$ is transformed using a sigmoid function to produce a conditional Bernoulli data likelihood over the image pixels $\oimage$ given the latent variables $\Lat$ and the input caption $\icaption$,
\begin{align}
P(\oimage\given\icaption, \Lat_{1:T}) = \prod_i P(x_i\given\icaption, \Lat_{1:T}) = \prod_i \sigmoid(\canv_{T,i}),
\end{align}
where, $\canv_{T,i}$ is the $i^{th}$ pixel on the final canvas computed at step T and $\sigmoid$ operator is a logistic sigmoid function $\sigmoid(x)=\frac{1}{1+exp(-x)}$. 
\comm{
We extended the architecture of the DRAW network to include an additional input caption from the language model described in Sec. (\ref{sec:lang}). Similarly to the original DRAW network, the conditional DRAW network is a stochastic recurrent neural network that consists of an Inference LSTM that infers the distribution of latent variables of image $x$ given caption $y$ and then a Generative LSTM that uses the inferred latent variables in order to reconstruct the image $x$ given caption $y$, where $\icaption$ is the input caption consisting of $N$ words $y_{1}, y_{2}, ..., y_{n}$. The $align$ function is used to compute the alignment between the input caption and intermediate image generative steps as in \cite{bahdanau_mt}:

Formally, the image is generated by iteratively computing the following equations for $t=1,...,T$
\begin{align}
\label{eq:x_hat}
\hat{x}_t &= x-\sigmoid(\canv_{t-1})\\
\label{eq:read}
r_t &= \readop(x_t, \hat{x}_t, \hdec_{t-1})\\
\henc_t &= \encoder(\henc_{t-1}, [r_t, \hdec_{t-1}])\\
\lat_t &\sim \post(\Lat_t|\henc_t)\\
\hdec_t &= \decoder(\hdec_{t-1}, z_t, s_{t-1})\\
s_{t} &= align(\hdec_{t-1}, \hlangall) \\
\label{eq:write}
\canv_t &= \canv_{t-1} + \writeop(\hdec_t)
\end{align}

where $\writeop$ is the same attention operators as in \citep{gregor_draw}.
%where $\readop$ and $\writeop$ are the same attention operators as in \citep{gregor_draw}.
% which apply a grid of Gaussian filters by specifying coordinates of the location and scale in order to yield a desirable image patch.
Given the caption representation from the language model, $\hlangall = [\hlang_{1}, \hlang_{2}, ..., \hlang_{N}]$, the $align$ operator computes the final sentence representation $s_t$ through a weighted sum using alignment probabilities $\alpha_{1...N}$:
\begin{align}
s_t=align(\hdec_{t-1}, \hlangall) = \alpha_{1}\hlang_{1} + \alpha_{2}\hlang_{2} + ... + \alpha_{N}\hlang_{N}.
\end{align}

The corresponding alignment probabilities $\alpha_{1...n}$ at each step are obtained by:
\begin{align}
e_{tj} &= v^{T}tanh(U\hlang_{j} + W\hdec_{t} + b)\\
\alpha_{j} &= \frac{exp(e_{tj})}{\sum_{j=1}^{N}exp(e_{tj})}.
\end{align}

Here $\hlang_{0}$ is initialized to the learned bias.
Setting $\alpha_{1...N}$ to $\frac{1}{N}$ turns the encoder into the vanilla model introduced in \citep{cho_mt} without the attention. 
}

@@

The terms in the variational lower bound are rearranged using the law of total expectation. Therefore, the loss function $\loss$ is calculated as follows:
\begin{align}
\loss =  &\expectation_{Q(\Lat_{1:T}\given\icaption,\oimage)}\left[ \log p(\oimage\given\icaption,\Lat_{1:T}) - \sum_{t=2}^T\kldiv\left(\post(\Lat_t\given\Lat_{t-1},\icaption,\oimage)\klBars\prior(\Lat_t\given\Lat_{t-1},\icaption)\right) \right] \nonumber \\& - \kldiv\left(\post(\Lat_1\given\oimage)\klBars\prior(\Lat_1)\right).
\end{align}
The expectation can be approximated by $\numSamples$ Monte Carlo samples $\LatSample_{1:T}$ from $\post(\Lat_{1:T}\given\icaption, \oimage)$:
\begin{align}
\loss \approx  &\frac{1}{\numSamples}\sum_{\sampleIdx=1}^\numSamples\left[ \log p(\oimage\given\icaption,\LatSample^\sampleIdx_{1:T}) - \sum_{t=2}^T\kldiv\left(\post(\Lat_t\given\LatSample^\sampleIdx_{t-1},\icaption,\oimage)\klBars\prior(\Lat_t\given\LatSample^\sampleIdx_{t-1},\icaption)\right) \right] \nonumber \\& - \kldiv\left(\post(\Lat_1\given\oimage)\klBars\prior(\Lat_1)\right).
\end{align}
We used only one sample during training for all our experiments.
\comm{
\begin{align}
\loss &= -\sum_{t=1}^{T}D_{KL}(\post(\Lat_t|\henc_t,s_{t-1})\,||\,\prior(\Lat_t)) + \frac{1}{L}\sum_{l=1}^{L}log\,p(x_{t}|y,z)\\
&=
\frac{1}{2}\sum_{t=1}^{T}(1 - 2\,log\,\sigma_{t}^{prior} + 2\,log\,\sigma_{t} - \frac{exp(2\,log\,\sigma_{t}) + (\mu_{t} - \mu_{t}^{prior})^{2}}{exp(2\,log\,\sigma_{t}^{prior})}) + \frac{1}{L}\sum_{l=1}^{L}log\,p(x_{t}|y,z)
\end{align}
}

@@

During the image generation step, we discard the inference network and instead sample from the prior distribution. 
Due to the blurriness of samples generated by the DRAW model, we perform an additional post processing step where we use a 
adversarial network trained on residuals of a Laplacian pyramid conditioned on the skipthought representation \citep{kiros_skipthought} of the captions 
to sharpen the generated images, similar to \citep{denton_lapgan}. By fixing the prior of the adversarial generator to its mean, it gets treated as a deterministic neural network which allows us to calculate the lower bound of likelihood. The reconstuction loss becomes the loss between sharpened image and correct image, whereas the latent loss stays the same. Additionally, we noticed that sampling from the mean of the uniform distribution generated samples with less noise than if we had otherwise sampled from the uniform distribution itself.

>>>>

Unlike the original DRAW network where latent variables are independent unit Gaussians $\mathcal{N}(0, I)$, the latent variables in the proposed alignDRAW model have their mean and variance depend on the previous recurrent hidden states $\hdec_{t-1}$, except for $\lat_1 \sim \prior(\Lat_1) = \mathcal{N}(0, I)$.  Namely, the mean and variance of prior distribution $\prior(\Lat_t)$ are parameterized by: 
\begin{align}
\mu_{t}(\hdec_{t-1}) &= tanh(W_{\mu}\hdec_{t-1}),\\
\sigma_{t}(\hdec_{t-1}) &= {exp}\big(tanh(W_{\sigma}\hdec_{t-1})\big),
\end{align}
where $W_{\mu}, W_{\sigma}$ are learnt model parameters. It has been shown in \citep{bachman_sdm} that model performance is improved by including dependencies between latent variables. An image is generated by iteratively computing the following equations for $t=1,...,T$ (see Figure~\ref{fig:figmodel}):
\begin{align}
\label{eq:x_hat}
\lat_t &\sim \prior(\Lat_t|\Lat_{1:t-1}) = \mathcal{N}\bigg(\mu_{t}(\hdec_{t-1}), \sigma_{t}(\hdec_{t-1})\bigg),\\
s_{t} &= align(\hdec_{t-1}, \hlangall), \\
\hdec_t &= \decoder(\hdec_{t-1}, [z_t, s_{t}]), \\
\canv_t &= \canv_{t-1} + \writeop(\hdec_t), \\
\tilde{\oimage} &\sim P(\oimage\given\icaption, \Lat_{1:T}) = \prod_i P(x_i\given\icaption, \Lat_{1:T}) = \prod_i \sigmoid(\canv_{T,i}),
\label{eq:write}
%\canv_t &= \canv_{t-1} + \writeop(\hdec_t)
\end{align}
where $\mathit{LSTM}$ is a standard \textit{Long Short-Term Memory} architecture \citep{hochreiter_lstm} with \textit{forget} gates \citep{gers_forget}, $[v_1, v_2]$ is the concatenation of the vectors $v_1$ and $v_2$ into the single vector, and $\canv_0$, $\hdec_0$ are initialized to the the learned biases. The $\writeop$ operator applies an array of Gaussian filters on the patch that places it on the canvas with smoothly varying location and zoom (same as in \citep{gregor_draw}). The $align$ function is used to compute the alignment between the input caption and intermediate image generative steps \citep{bahdanau_mt}. Given the caption representation from the language model, $\hlangall = [\hlang_{1}, \hlang_{2}, ..., \hlang_{N}]$, the $align$ operator outputs a dynamic sentence representation $s_t$ at each step through a weighted sum using alignment probabilities $\alpha_{1...N}^{t}$:
%\begin{align}
%\vspace{-0.1in}
$s_t=align(\hdec_{t-1}, \hlangall) = \alpha_{1}^{t}\hlang_{1} + \alpha_{2}^{t}\hlang_{2} + ... + \alpha_{N}^{t}\hlang_{N}.$
%\end{align}
The corresponding alignment probabilities $\alpha_{1...N}^{t}$ at each step are obtained using the caption representation $\hlangall$ and the hidden state of the generative model $\hdec_{t}$, by first calculating energies using a $tanh$ perceptron scaled by vector $v$ and then normalizing them, as in \citep{bahdanau_mt}:
%\begin{align}
%\vspace{-0.1in}
$\alpha_{k}^{t}=\frac{exp(v^{T}tanh(U\hlang_{k} + W\hdec_{t} + b)}{\sum_{j=1}^{N}exp(v^{T}tanh(U\hlang_{j} + W\hdec_{t} + b)}.$
%\end{align}
The $i^{th}$ pixel $\canv_{T,i}$ from the final canvas $\canv_T$ is transformed using a sigmoid function $\sigmoid$ to produce a conditional Bernoulli data likelihood over the image pixels $\oimage$ given the latent variables $\Lat$ and the input caption $\icaption$,

@@

The terms in the variational lower bound are rearranged using the law of total expectation. Therefore, the objective function $\loss$ is calculated as follows:
\begin{align}
\label{eq:loss}
\loss =  &\expectation_{Q(\Lat_{1:T}\given\icaption,\oimage)}\left[ \log p(\oimage\given\icaption,\Lat_{1:T}) - \sum_{t=2}^T\kldiv\left(\post(\Lat_t\given\Lat_{t-1},\icaption,\oimage)\klBars\prior(\Lat_t\given\Lat_{t-1},\icaption)\right) \right] \nonumber \\& - \kldiv\left(\post(\Lat_1\given\oimage)\klBars\prior(\Lat_1)\right).
\end{align}
The expectation can be approximated by $\numSamples$ Monte Carlo samples $\LatSample_{1:T}$ from $\post(\Lat_{1:T}\given\icaption, \oimage)$:
\begin{align}
\label{eq:loss_mc}
\loss \approx  &\frac{1}{\numSamples}\sum_{\sampleIdx=1}^\numSamples\left[ \log p(\oimage\given\icaption,\LatSample^\sampleIdx_{1:T}) - \sum_{t=2}^T\kldiv\left(\post(\Lat_t\given\LatSample^\sampleIdx_{t-1},\icaption,\oimage)\klBars\prior(\Lat_t\given\LatSample^\sampleIdx_{t-1},\icaption)\right) \right] \nonumber \\& - \kldiv\left(\post(\Lat_1\given\oimage)\klBars\prior(\Lat_1)\right).
\end{align}
We used only one sample during training for all our experiments.
\comm{
\begin{align}
\loss &= -\sum_{t=1}^{T}D_{KL}(\post(\Lat_t|\henc_t,s_{t-1})\,||\,\prior(\Lat_t)) + \frac{1}{L}\sum_{l=1}^{L}log\,p(x_{t}|y,z)\\
&=
\frac{1}{2}\sum_{t=1}^{T}(1 - 2\,log\,\sigma_{t}^{prior} + 2\,log\,\sigma_{t} - \frac{exp(2\,log\,\sigma_{t}) + (\mu_{t} - \mu_{t}^{prior})^{2}}{exp(2\,log\,\sigma_{t}^{prior})}) + \frac{1}{L}\sum_{l=1}^{L}log\,p(x_{t}|y,z)
\end{align}
}

@@

During the image generation step, we discard the inference network and instead sample from the prior distribution. 
Due to the blurriness of samples generated by the DRAW model, we perform an additional post processing step where we use a 
adversarial network trained on residuals of a Laplacian pyramid conditioned on the skipthought representation \citep{kiros_skipthought} of the captions 
to sharpen the generated images, similar to \citep{denton_lapgan}. By fixing the prior of the adversarial generator to its mean, it gets treated as a deterministic neural network that allows us to define the conditional data term in Eq. (\ref{eq:loss}) on the sharpened images and calculate the lower bound accordingly. Additionally, we noticed that sampling from the mean of the uniform distribution generated samples with less noise than if we had otherwise sampled from the uniform distribution itself.
